\chapter{Hardware Requirements}
\label{app:hardware}

This appendix provides concrete hardware requirements and recommendations for
training and running MicroGPT.  All estimates assume the default configuration:
135M parameters, 1\,024 context length, 32K vocabulary, bf16 mixed precision.


\section{Memory Breakdown}

The GPU memory footprint has four components:

\begin{center}
\begin{tabular}{lrl}
\toprule
Component & Size & Notes \\
\midrule
Model weights (fp32 master) & 0.54 GB & $135\text{M} \times 4$ bytes \\
AdamW optimizer states ($m$, $v$) & 1.08 GB & $2 \times 135\text{M} \times 4$ bytes \\
Gradients (fp32) & 0.54 GB & $135\text{M} \times 4$ bytes \\
Activations (bf16, batch=16) & $\sim$6 GB & Scales with batch size \\
\midrule
\textbf{Total (training, batch=16)} & $\sim$\textbf{8.2 GB} & \\
\bottomrule
\end{tabular}
\end{center}

Activations dominate and scale linearly with batch size.  Reducing
\code{DEVICE\_BATCH\_SIZE} from 16 to 8 cuts activation memory roughly in half;
gradient accumulation compensates to maintain the same effective batch size.

For \textbf{inference only}, the footprint is much smaller --- the model in bf16
plus the KV cache:

\begin{center}
\begin{tabular}{lrl}
\toprule
Component & Size & Notes \\
\midrule
Model weights (bf16) & 0.27 GB & $135\text{M} \times 2$ bytes \\
KV cache (12 layers, bf16) & 0.04 GB & $12 \times 2 \times 1024 \times 12 \times 64 \times 2$ bytes \\
\midrule
\textbf{Total (inference)} & $\sim$\textbf{0.3 GB} & \\
\bottomrule
\end{tabular}
\end{center}


\section{Dataset Storage}

FineWeb-Edu shards are compressed Parquet files hosted on HuggingFace:

\begin{center}
\begin{tabular}{rrrl}
\toprule
Shards & Disk & $\sim$Tokens & Use case \\
\midrule
2 & $\sim$150 MB & $\sim$200M & Smoke test / debugging \\
10 (default) & $\sim$700 MB & $\sim$1B & Quick experiment \\
50 & $\sim$3.5 GB & $\sim$5B & Solid training run \\
99 (all) & $\sim$7 GB & $\sim$10B & Full reproduction \\
\bottomrule
\end{tabular}
\end{center}

The last shard is reserved for validation.  For the default 20K-step run at
0.5M tokens/step ($\sim$10B tokens total), download all 99 shards.  With fewer
shards the dataloader loops (multi-epoch), which still works but may overfit.

Checkpoint storage is modest: each checkpoint is $\sim$1.1 GB (model + optimizer
state in fp32).


\section{Training: GPU Recommendations}

Training time depends on GPU throughput.  The estimates below assume 40\%
Model FLOPs Utilization (MFU), which is realistic for this model size with
\code{torch.compile} and bf16:

\begin{center}
\begin{tabular}{lrrrr}
\toprule
GPU & VRAM & bf16 TFLOPS & 1-GPU time & 8-GPU time \\
\midrule
RTX 3080       & 10 GB & 30  & $\sim$200 h  & $\sim$25 h \\
RTX 3090       & 24 GB & 36  & $\sim$166 h  & $\sim$21 h \\
RTX 4090       & 24 GB & 83  & $\sim$72 h   & $\sim$9 h \\
A100 (40/80 GB) & 40--80 GB & 312 & $\sim$19 h   & $\sim$2.4 h \\
H100 SXM       & 80 GB & 990 & $\sim$6 h    & $\sim$45 min \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Minimum}

A \textbf{single GPU with 10 GB VRAM} (e.g.\ RTX 3080) can train MicroGPT.
Reduce \code{DEVICE\_BATCH\_SIZE} to 4--8 to fit in memory; gradient accumulation
will maintain the effective batch size.  Training will take roughly a week.

\subsection{Recommended}

\begin{itemize}
  \item \textbf{Budget option}: 1$\times$ RTX 4090 on vast.ai ($\sim$\$0.40/hr).
    Finishes in $\sim$3 days, total cost $\sim$\$35.
  \item \textbf{Sweet spot}: 1$\times$ A100 80\,GB ($\sim$\$1.50/hr).  Comfortable
    VRAM headroom, finishes in under a day, $\sim$\$30.
  \item \textbf{Fast iteration}: 8$\times$ H100 node ($\sim$\$25/hr on vast.ai).
    Under an hour for a full run, $\sim$\$25.  Matches nanochat's reported
    training time for GPT-2 reproduction.
\end{itemize}

\subsection{Batch size adjustments}

If you hit out-of-memory errors, reduce \code{DEVICE\_BATCH\_SIZE} in
\code{microgpt/train.py}.  The gradient accumulation logic automatically
increases \code{grad\_accum\_steps} to compensate:

\[
  A = \left\lfloor \frac{B_{\text{total\_tokens}}}{B_{\text{device}} \times T \times W} \right\rfloor
\]

\begin{center}
\begin{tabular}{rrrl}
\toprule
\code{DEVICE\_BATCH\_SIZE} & Peak VRAM & Accum steps (1 GPU) & Notes \\
\midrule
16 & $\sim$8.2 GB & 32 & Default \\
8 & $\sim$5.1 GB & 64 & Fits on 8 GB GPUs \\
4 & $\sim$3.6 GB & 128 & Fits on 6 GB GPUs \\
\bottomrule
\end{tabular}
\end{center}

Smaller micro-batches increase the number of accumulation steps per optimizer
step, which is slightly slower (more kernel launches) but mathematically
equivalent.


\section{Inference and Evaluation}

Inference (text generation) and evaluation (computing validation loss / BPB)
have much lower requirements than training:

\begin{center}
\begin{tabular}{lrl}
\toprule
Task & Min VRAM & Notes \\
\midrule
Text generation (batch=1) & $\sim$0.5 GB & Model (bf16) + KV cache \\
Evaluation (\code{microgpt.eval}, batch=8) & $\sim$3 GB & No optimizer states \\
\bottomrule
\end{tabular}
\end{center}

\textbf{Any modern GPU with 2+ GB VRAM} can run inference.  This includes
laptop GPUs (e.g.\ RTX 3050 Mobile with 4 GB).  CPU-only inference also works
but is $\sim$50$\times$ slower.

For evaluation, a single GPU with 4+ GB VRAM is sufficient.  Reduce
\code{-{}-batch-size} if needed.

\tip{If you trained on a multi-GPU cluster but want to evaluate on a single
consumer GPU, just load the checkpoint --- the model weights are the same
regardless of how many GPUs were used for training.  DDP only affects
gradient averaging, not the final model.}


\section{Summary}

\begin{center}
\begin{tabular}{lll}
\toprule
Task & Minimum & Recommended \\
\midrule
Training & 1$\times$ GPU, 10 GB VRAM & 1$\times$ A100 80\,GB or 8$\times$ H100 \\
Inference & Any GPU, 2 GB VRAM & Any GPU, 4+ GB VRAM \\
Evaluation & 1$\times$ GPU, 4 GB VRAM & 1$\times$ GPU, 8+ GB VRAM \\
Disk (dataset) & 150 MB (2 shards) & 7 GB (all 99 shards) \\
Disk (checkpoints) & 1.1 GB & 5+ GB (multiple saves) \\
\bottomrule
\end{tabular}
\end{center}
